DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-10-05 Paper_May_2012/main.tex
Paper: Anmerkungen von Winni eingearbeitet sowie Referenzen verlinkt und nicht referenzierte labels entfernt

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
%\todo{Aus der Sicht des Users anhand eines konkreten Beispiels beschreiben}
%\todo{Einfaches Beispiel nehmen}
%\todo{erwaehnen, dass wir alles immer nicht-linear betrachten}
%\todo{Aussehen eines typischen PDE und Opt loop und deren Gemeinsamkeiten
%erklaeren. Vielleicht mir Pseudocode}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). 
\subsection{Solving of PDEs}
\subsubsection{Stationary problems}
The easiest case 
is that of a PDE in weak form to find some $u$
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$. As example, we take 
the Laplace problem in weak form:
\[
a(u)(\phi) = (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
To solve the problem, we need a configuration (a mesh), eventually 
a diffusion parameter, a finite element, a quadrature rule,
a linear solver (here the CG solver), and some output such as
a graphical solution and eventually some measurement functional.
Each of those routines are written in the \texttt{main.cc} file. 
We start by including necessary files both the DOpElib library and
the deal.II software:
\begin{lstlisting}
  #include "pdeproblemcontainer.h" etc. ...
\end{lstlisting}
In the \texttt{main} function, the following steps need to be done. First, 
the parameter reader for runtime parameters is initialized:
\begin{lstlisting}
  ParameterReader pr;
  SSolver::declare_params(pr);
  DOpEOutputHandler<VECTOR>::declare_params(pr);
  pr.read_parameters(paramfile);
\end{lstlisting}
A linear finite element in two dimensions and two components is 
initialized:
\begin{lstlisting}
  FESystem<2> state_fe(FE_Q<2>(1), 2);
\end{lstlisting}
and a quadrature rule of order three is used to approximate the 
integrals:
\begin{lstlisting}
  QGauss<2> quadrature_formula(3);
\end{lstlisting}
Then, the \texttt{LocalPDE} object comes into play. This is an quite 
important feature of DOpElib because it allows for easy implementation 
of the variational formulation. Specific explication is given below because
the implementation is seperated in \texttt{localpde.h}.
Afterwards, a \texttt{LocalPointFunctional} is declared. As for the PDE, the 
implementation is done in another file, namely \texttt{localfunctional.h}.

Generically, each problem in DOpElib is considered to be time-dependent and nonlinear because
these are the 'real' world problems. Therefore, (even for the linear and stationary Laplace equation),
we have to declare a (pseudo) time:
\begin{lstlisting}
  std::vector<double> times(1, 0.);
\end{lstlisting}
Then, the spatial mesh (here the unit square) is created and three-times globally-refined:
\begin{lstlisting}
  GridGenerator::hyper_cube(triangulation, 0, 1);
  triangulation.refine_global(3);
\end{lstlisting}
A major component which organizes the whole solution process (independent from 
the problem!) is the 
\begin{lstlisting}
  MethodOfLines_StateSpaceTimeHandler<FE, DOFHANDLER, 
    SPARSITYPATTERN, VECTOR, 2> DOFH(triangulation, state_fe);
\end{lstlisting}
This method handles all dofs in space and time and different versions exist
for forward as well as optimization problems.

@@

  P.SetDirichletBoundaryColors(0, comp_mask, &DD1);
\end{lstlisting}
Next, the solver for the linear equations needs to be specified:
\begin{lstlisting}
  SSolver solver(&P, "fullmem", pr, idc);
\end{lstlisting}
Finally, the ouput handler is initialized via
\begin{lstlisting}
  P.RegisterOutputHandler(&out);
  solver.RegisterOutputHandler(&out);
\end{lstlisting}
Finally, the programm is started with:
\begin{lstlisting}
  solver.ReInit();
  out.ReInit();
  out.Write(outp, 1, 1, 1);
  solver.ComputeReducedFunctionals();
\end{lstlisting}
After explaining the basic compononents that are necessary for all  
(linear and stationary) partial differential equations, we shall give 
an overview to the implementation of the specific form of the PDE. 
This step is done in the \texttt{localpde.h} file. As already explained,
each PDE is treatet in a nonlinear manner such that for solving 
the Laplace problem the following nonlinear step is necessary to recapitulate:
Find $\delta u\in V$:
\[
(\nabla \delta u, \nabla \phi) = -(\nabla u, \nabla \phi) 
\]
and $u^{n+1} = u^n + \delta u$ for $n=1,2,3,ldots$.
Therefore, the implementation reads for the right hand side:
\begin{lstlisting}
   void CellEquation(...)
    {
      // implement (\nabla u, \nabla \phi) 
    }
\end{lstlisting}
Specifically, the inner loops reads
\begin{lstlisting}
   for (all quadrature points per cell q)
      {
        Tensor<2, 2> ugrads; // solution vector

@@

          local_cell_vector(i) += scale * 
            scalar_product(ugrads, phi_i_grads_u) *
	    state_fe_values.JxW(q_point);
        }
      }
\end{lstlisting}
For the left hand side, we need to implement the Jacobian matrix:
\begin{lstlisting}
   void CellMatrix(...)
    {
      // implement  (\nabla \delta u, \nabla \phi)
    }
\end{lstlisting}
Specifically,
\begin{lstlisting}
      for (all quadrature points per cell q)      
        for (all dofs per cell i)    
          for (all dofs per cell j)
          {
            local_entry_matrix(i, j) += 
              scalar_product(phi_grads_u[j], phi_grads_u[i]) * 
              state_fe_values.JxW(q_point);
          }
\end{lstlisting}
The integration of the equation for each DoF and each quadrature point on each
cell is then done 
in the \texttt{templates/integrator.h} function.

@@

\subsection{Solving of optimization problems}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
As before, the main part is done in the \texttt{main} function there is no
great difference to the previous explications. However, 
since we utilize gradient-based optimization routines \cite{...},
all derivatives must be implemented. All this information
must given to the \texttt{localpde.h} file. Herein, one has to 
implement the adjoint, tangent and adjoint Hession equations. 
The algorithmic parts are described in the literature \cite{BeMeVe06}.

@@

Minimize the cost functional $J(q,U)$ subject to the state equation
$A(q,U)(\Phi) = 0$ (as defined
in~\eqref{state_form} and~\eqref{opt_state_equation}) for $(q,U) \in 

@@

@@

will be specified in Algorithm \ref{algo_opt}.

@@

Let $q\in Q_d, U=S(q)\in {\cal X}$
and $Z\in {\cal X}$ the dual solution be obtained after solving 

@@

j'(q)(\delta q) := {\cal L}'_{q}(q,U,Z)(\delta q),

@@

A_{q}'(q,U)(\tau q , Z).

@@

  Let $q\in Q_d, U=S(q)\in {\cal X}$ and $Z\in {\cal X}$ and  $\delta
  q\in Q_d$ be given. Further, let $\delta U\in {\cal X}$ be the

@@

  \delta U\in{\cal X},\quad 
  A_{U}'(q,U)(\delta U , \Phi) 
  = -A_{q}'(q,U)(\delta q , \Phi)  \quad\forall \Phi\in {\cal

@@

  and $\delta Z\in {\cal X}$ the solution of the adjoint Hessian 

@@

  A_{U}'(q,U)(\Phi, \delta Z)
  = J''_{UU}(q,U)(\delta U, \Phi)
  - A_{UU}''(q,U)(\delta U, \Phi, Z)
  - A_{qU}''(q,U)(\delta q , \Phi, Z)  \quad\forall \Phi\in

@@

  A_{qq}''(q, U)(\delta q, \tau q , Z) -
  A_{Uq}''(q,U)(\delta U, \tau q , Z) -
  A_{q}'(q,U)(\tau q , \delta Z).

@@

    A_{U}'(q,U)(\Phi , Z) &= J'_{U}(q,U)(\Phi) & \forall \Phi&\in {\cal X},

@@

The configuration is 
sketched in Figure \ref{configuration_csm_and_fsi_2D}. Details 
on parameters and evaluation functionals and other results 
can be found in \cite{HrTu06b, BuSc06, TuHrMaRaWoAc10, DeHaeAnnBrVie10, Wi11}. 
The Fractional-Step-$\theta$ scheme is used for time discretization with
different time step sizes $k$. However, the time stepping scheme can be 
very easily chosen in the \texttt{main.cc} function by declaring
\begin{lstlisting}
#define TSP1 ForwardEulerProblem
#define TSP2 BackwardEulerProblem
#define TSP3 CrankNicolsonProblem
#define TSP4 ShiftedCrankNicolsonProblem
#define TSP5 FractionalStepThetaProblem
\end{lstlisting}
and then using in the code the designated scheme, 
for example here the Fractional-Step-$\theta$ scheme
\texttt{TSP5}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
%\todo{Aus der Sicht des Users anhand eines konkreten Beispiels beschreiben}
%\todo{Einfaches Beispiel nehmen}
%\todo{erwaehnen, dass wir alles immer nicht-linear betrachten}
%\todo{Aussehen eines typischen PDE und Opt loop und deren Gemeinsamkeiten
%erklaeren. Vielleicht mir Pseudocode}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). 
\subsection{Solving of PDEs}
\subsubsection{Stationary problems}
The easiest case 
is that of a PDE in weak form to find some $u$
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$. As example, we take 
the Laplace problem in weak form:
\[
a(u)(\phi) = (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
To solve the problem, we need a configuration (a mesh), eventually 
a diffusion parameter, a finite element, a quadrature rule,
a linear solver (here the CG solver), and some output such as
a graphical solution and eventually some measurement functional.
Each of those routines are written in the \texttt{main.cc} file. 
We start by including necessary files both the DOpElib library and
the deal.II software:
\begin{lstlisting}
  #include "pdeproblemcontainer.h" etc. ...
\end{lstlisting}
In the \texttt{main} function, the following steps need to be done. First, 
the parameter reader for runtime parameters might be initialized. In a parameter
file the user might specifiy material paramters for the concrete PDE, but also
general parameters for nonlinear and linear solvers, and output-related 
aspects.

A linear finite element in two dimensions and two components is 
initialized:
\begin{lstlisting}
  FESystem<2> state_fe(FE_Q<2>(1), 2);
\end{lstlisting}
and a quadrature rule of order three is used to approximate the 
integrals:
\begin{lstlisting}
  QGauss<2> quadrature_formula(3);
\end{lstlisting}
Then, the \texttt{LocalPDE} object comes into play. This is an quite 
important feature of DOpElib because it allows for easy implementation 
of the variational formulation. Specific explication is given below because
the implementation is seperated in \texttt{localpde.h}.
Afterwards, a \texttt{LocalPointFunctional} is declared. As for the PDE, the 
implementation is done in another file, namely \texttt{localfunctional.h}.

Generically, each problem in DOpElib is considered to be time-dependent and nonlinear because
these are the 'real' world problems. Therefore, (even for the linear and stationary Laplace equation),
we have to declare a (pseudo) time:
\begin{lstlisting}
  std::vector<double> times(1, 0.);
\end{lstlisting}
Then, the spatial mesh (here the unit square) is created and three-times globally-refined:
\begin{lstlisting}
  GridGenerator::hyper_cube(triangulation, 0, 1);
  triangulation.refine_global(3);
\end{lstlisting}
A major component which organizes the whole solution process (independent from 
the problem!) is the 
\begin{lstlisting}
  MethodOfLines_StateSpaceTimeHandler<...> DOFH (...);
\end{lstlisting}
This method handles all dofs in space and time and different versions exist
for forward as well as optimization problems.

@@

  P.SetDirichletBoundaryColors(0, comp_mask, &DD1);
\end{lstlisting}
Next, the solver for the linear equations needs to be specified:
\begin{lstlisting}
  SSolver solver(&P, "fullmem", pr, idc);
\end{lstlisting}
Finally, the ouput handler is initialized via
\begin{lstlisting}
  P.RegisterOutputHandler(&out);
  solver.RegisterOutputHandler(&out);
\end{lstlisting}
Finally, the programm is started with:
\begin{lstlisting}
  solver.ReInit();
  out.ReInit();
  out.Write(outp, 1, 1, 1);
  solver.ComputeReducedFunctionals();
\end{lstlisting}
After explaining the basic compononents that are necessary (and always the same in DOpElib) 
for all  (linear and stationary) partial differential equations, we shall give 
an overview to the implementation of the specific form of the PDE. 
Beside specification of boundary conditions, the form of the local PDE depends 
on the concrete problem under consideration. 
This step is done in the \texttt{localpde.h} file. As already explained,
each PDE is treatet in a nonlinear manner such that for solving 
the Laplace problem the following nonlinear step is necessary to recapitulate:
Find $\delta u\in V$:
\[
(\nabla \delta u, \nabla \phi) = -(\nabla u, \nabla \phi) 
\]
and $u^{n+1} = u^n + \delta u$ for $n=1,2,3,ldots$.
Therefore, the implementation reads for the right hand side:
\begin{lstlisting}
   void CellEquation(...)
    {
      // implement (\nabla u, \nabla \phi) 
    }
\end{lstlisting}
Specifically, the inner loops reads
\begin{lstlisting}
   for (all quadrature points per cell q)
      {
        Tensor<2, 2> ugrads; // solution vector

@@

          local_cell_vector(i) += scale * 
            scalar_product(ugrads, phi_i_grads_u) *
	    state_fe_values.JxW(q_point);
        }
      }
\end{lstlisting}
For the left hand side, we need to implement the Jacobian matrix:
\begin{lstlisting}
   void CellMatrix(...)
    {
      // implement  (\nabla \delta u, \nabla \phi)
    }
\end{lstlisting}
Specifically,
\begin{lstlisting}
      for (all quadrature points per cell q)      
        for (all dofs per cell i)    
          for (all dofs per cell j)
          {
            local_entry_matrix(i, j) += 
              scalar_product(phi_grads_u[j], phi_grads_u[i]) * 
              state_fe_values.JxW(q_point);
          }
\end{lstlisting}
The integration of the equation for each DoF and each quadrature point on each
cell is then done 
in the \texttt{templates/integrator.h} function. Finally, as usually in 
finite element computations, this leads to a discrete system. Since
we treat each problem with a Newton method, within each Newton step, the 
following linear system must be solved:
\[
A\delta u = b
\]
in which $A$ is specified by \texttt{CellMatrix(...)} and 
$b$ by \texttt{CellEquation(...)}. The vector $\delta u$ denotes the solution to the 
present Newton step:
\[
u^n = u^{n-1} + \omega \delta u
\]
with a line search parameter $\omega\in (0,1]$.

@@

\subsection{Solving of optimization problems}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
As before, the main part is done in the \texttt{main} function there is no
great difference to the previous explications. However, 
since we utilize gradient-based optimization routines in the philosophy of
\cite{Troe99,HiKu01},
all derivatives must be implemented. All this information
must given to the \texttt{localpde.h} file. Herein, one has to 
implement the adjoint, tangent and adjoint Hession equations. 
The algorithmic parts are described in the literature \cite{BeMeVe06}.

@@

Minimize the cost functional $J(q,u)$ subject to the state equation
$a(q,u)(\phi) = 0$ for $(q,u) \in 

@@

Thus, the constrained optimization problem can be formulated by means
of  
\begin{defi}[Unconstrained optimization]
  \label{opt_reduced_problem}
  Minimize $j(q)$ for $q\in Q_d$.

@@

will are implemented in \texttt{reducednewtonalgorithm.h}.

@@

Let $q\in Q_d, u=S(q)\in {\cal X}$
and $z\in {\cal X}$ the dual solution be obtained after solving 

@@

j'(q)(\delta q) := {\cal L}'_{q}(q,u,z)(\delta q),

@@

A_{q}'(q,u)(\tau q , z).

@@

  Let $q\in Q_d, u=S(q)\in {\cal X}$ and $z\in {\cal X}$ and  $\delta
  q\in Q_d$ be given. Further, let $\delta u\in {\cal X}$ be the

@@

  \delta u\in{\cal X},\quad 
  a_{u}'(q,u)(\delta u , \phi) 
  = -a_{q}'(q,u)(\delta q , \phi)  \quad\forall \phi\in {\cal

@@

  and $\delta z\in {\cal X}$ the solution of the adjoint Hessian 

@@

  a_{u}'(q,u)(\phi, \delta z)
  = J''_{uu}(q,u)(\delta u, \phi)
  - a_{uu}''(q,u)(\delta u, \phi, z)
  - a_{qu}''(q,u)(\delta q , \phi, z)  \quad\forall \phi\in

@@

  a_{qq}''(q, u)(\delta q, \tau q , z) -
  a_{uq}''(q,u)(\delta u, \tau q , z) -
  a_{q}'(q,u)(\tau q , \delta z).

@@

    a_{u}'(q,u)(\phi , z) &= J'_{u}(q,u)(\phi) & \forall \phi&\in {\cal X},

@@

The configuration is 
sketched in Figure \ref{configuration_csm_and_fsi_2D}. 
The computational domain has length $L=2.5m$ and height $H=0.41m$. The circle center
is positioned at $C=(0.2m,0.2m)$ with radius $r=0.05m$. The elastic beam has length
$l=0.35m$ and height $h=0.02m$. The right lower end is positioned at 
$(0.6m,0.19m)$, and
the left end is attached to the circle.

\begin{figure}[h]
\centering
\input{Pictures/pic_12}
\caption{Flow around cylinder with elastic beam with 
circle-center $C=(0.2,0.2)$ and radius $r=0.05$.}
\label{configuration_csm_and_fsi_2D}
\end{figure}

Control points $A(t)$ (with $A(0) = (0.6,0.2)$) are fixed at the 
trailing edge of the structure, measuring $x$- and $y$-deflections of the beam.

Details 
on parameters and evaluation functionals and other results 
can be found in \cite{HrTu06b,BuSc06,DeHaeAnnBrVie10,Wi11}. 
The Fractional-Step-$\theta$ scheme is used for time discretization with
different time step sizes $k$. However, the time stepping scheme can be 
very easily chosen in the \texttt{main.cc} function by declaring
\begin{lstlisting}
#define TSP1 ForwardEulerProblem
#define TSP2 BackwardEulerProblem
#define TSP3 CrankNicolsonProblem
#define TSP4 ShiftedCrankNicolsonProblem
#define TSP5 FractionalStepThetaProblem
\end{lstlisting}
and then using in the code the designated scheme, 
for example here the Fractional-Step-$\theta$ scheme
\texttt{TSP5}.
